---
title: 'Gut Microbiome Analysis: Clean'
author: Benjamin Marks and Angie Zhu
output: html_document
date: "2025-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code created by Benjamin Marks and Angie Zhu

## 0 - Load Libraries
```{r libraries}

library(phyloseq)
library(vegan)
library(rstatix)
library(Maaslin2)
library(haven)    
library(ggrepel)
library(RColorBrewer)
library(rlang)
library(dplyr)
library(ggsci)
library(ggpubr)
library(patchwork)
library(viridis)
library(microbiomeMarker)
library(MicrobiotaProcess)
library(ggplot2)
library(ALDEx2)
```


## 1 - Load data
```{r load data}

ps_gutMicro <- readRDS("ben_phyloseq.rds") # put your filename here


# Clean taxonomy levels (if not done already)
tax_table_ps <- phyloseq::tax_table(ps_gutMicro)
colnames(tax_table_ps) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
tax_table(ps_gutMicro) <- tax_table_ps
```


## 2 - Select Variables
```{r variables}
# Access sample metadata, look at it and choose the variables you are interested in
meta <- as.data.frame(phyloseq::sample_data(ps_gutMicro)) # should have 57 samples/rows

# Investigating variables
variables1 <- c("EH", "GL", "C", "EH_GL", "EH_C", "GL_C", "EH_GL_C", "ANY")


meta$sex <- factor(
  meta$sex,
  levels = c(1, 2),        
  labels = c("Male", "Female")
) 
meta$EH <- factor(
  meta$EH,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 
meta$GL <- factor(
  meta$GL,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$C <- factor(
  meta$C,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$EH_GL <- factor(
  meta$EH_GL,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$EH_C <- factor(
  meta$EH_C,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$GL_C <- factor(
  meta$GL_C,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$EH_GL_C <- factor(
  meta$EH_GL_C,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 

meta$ANY <- factor(
  meta$ANY,
  levels = c(0, 1),        
  labels = c("neg", "pos")
) 
phyloseq::sample_data(ps_gutMicro) <- meta
```


## 3 - Alpha Diversity Plot (Chao1 and Shannon index)
```{r alpha diversity}

# for categorical variables (box plot)
a_diversity_categorical <- function(ps, variable, title=variable) {
  
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  value_counts <- table(sam_data_filtered[[variable]])
  valid_values <- names(value_counts)[value_counts >= 1]
  final_keep_samples <- sam_data_filtered[[variable]] %in% valid_values
  ps_final <- prune_samples(final_keep_samples, ps_filtered)
  alpha_div <- estimate_richness(ps_final, measures = c("Chao1", "Shannon"))
  alpha_div$sample_id <- rownames(alpha_div)
  alpha_div <- merge(alpha_div, sam_data_filtered, by.x = "sample_id", by.y = "row.names")
  
  phyloseq::sample_data(ps_final)[[variable]] <- factor(phyloseq::sample_data(ps_final)[[variable]])
  
  a_diversity_factor <- plot_richness(ps_final, x = variable, color = variable, 
                                      measures = c("Chao1", "Shannon"))
  a_diversity_factor$layers[[2]] = NULL 
  
  a_diversity_factor <- a_diversity_factor + 
    geom_boxplot() + 
    geom_jitter(alpha = 0.25) + 
    theme_bw() + 
    theme(axis.text = element_text(size = 10),
          axis.title = element_text(size = 12, face = "bold"), 
          legend.position = "none", 
          axis.text.x = element_text(size = 10)) + 
    labs(x = title) + 
    stat_compare_means(
      method = "kruskal.test",
      label = "p.format",
      label.x = 1,
      label.y = Inf,
      hjust = 1.1,
      vjust = 1.5,
      size = 3.5
    )
  
  print(a_diversity_factor)
  return(a_diversity_factor)
}


# for continuous variables (scatter plot)
a_diversity_continuous <- function(ps, variable, title = variable, measures = c("Chao1", "Shannon")) {
  
  # Estimate richness
  alpha_div <- estimate_richness(ps, measures = measures)
  rownames(alpha_div) <- sub("^X", "", rownames(alpha_div))
  
  sdf <- as.data.frame(phyloseq::sample_data(ps))
  variable <- as.character(variable)
  if (!(variable %in% colnames(sdf))) {
    stop(paste("Variable", variable, "not found in phyloseq::sample_data!"))
  }
  tmp_var <- sdf[rownames(alpha_div), variable]
  alpha_div[[variable]] <- as.numeric(unlist(tmp_var))
  alpha_div <- alpha_div[!is.na(alpha_div[[variable]]), ]
  
  # Left plot: y-axis label, no x-axis label
  p1 <- ggscatter(
    alpha_div, x = variable, y = measures[1],
    add = "reg.line", conf.int = TRUE,
    cor.coef = TRUE, cor.method = "pearson",
    xlab = NULL, ylab = "Alpha Diversity Measure",
    size = 2
  ) +
    theme_bw() +
    theme(
      axis.text = element_text(size = 12),
      axis.title.y = element_text(size = 13, face = "bold", angle = 90, vjust = 1),
      axis.title.x = element_blank(),
      plot.title = element_text(hjust = 0.5)
    ) +
    ggtitle(measures[1])

  # Right plot: no axis labels
  p2 <- ggscatter(
    alpha_div, x = variable, y = measures[2],
    add = "reg.line", conf.int = TRUE,
    cor.coef = TRUE, cor.method = "pearson",
    xlab = NULL, ylab = NULL,
    size = 2
  ) +
    theme_bw() +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_blank(),
      plot.title = element_text(hjust = 0.5)
    ) +
    ggtitle(measures[2])

  # Combine with patchwork
  combined_plot <- p1 + p2 +
    plot_layout(nrow = 1, guides = "collect") &
    theme(
      axis.title.x = element_blank() # keep blank!
    )

  # Add x-axis label below using patchwork::wrap_elements
  combined_plot <- combined_plot / 
    patchwork::wrap_elements(
      grid::textGrob(title, gp = grid::gpar(fontsize = 13, fontface = "bold"))
    ) +
    plot_layout(heights = c(10, 1))

  print(combined_plot)
  return(combined_plot)
}
EH <- a_diversity_categorical(ps_gutMicro, variable = "EH", title = "Entamoeba histolytica")
GL <- a_diversity_categorical(ps_gutMicro, variable = "GL", title = "Giardia lamblia")
C <- a_diversity_categorical(ps_gutMicro, variable = "C", title = "Cryptosporidium")

ggsave("EH_alpha.png", plot = EH, width = 6, height = 4, dpi = 500)
ggsave("GL_alpha.png", plot = GL, width = 6, height = 4, dpi = 500)
ggsave("C_alpha.png", plot = C, width = 6, height = 4, dpi = 500)
ggsave("ANY_alpha.png", plot = ANY, width = 6, height = 4, dpi = 500)

ggsave("EH_GL_alpha.png", plot = EH_GL, width = 6, height = 4, dpi = 500)
ggsave("EH_C_alpha.png", plot = EH_C, width = 6, height = 4, dpi = 500)
ggsave("GL_C_alpha.png", plot = GL_C, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C_alpha.png", plot = EH_GL_C, width = 6, height = 4, dpi = 500)


EH_GL <- a_diversity_categorical(ps_gutMicro, variable = "EH_GL", title = "Entamoeba histloytica and Giardia Lamblia Coinfection")
EH_C <- a_diversity_categorical(ps_gutMicro, variable = "EH_C", title = "Entamoeba histloytica and Cryptosporidium Coinfection")
GL_C <- a_diversity_categorical(ps_gutMicro, variable = "GL_C", title = "Giardia lamblia and Cryptosporidium Coinfection")
EH_GL_C <- a_diversity_categorical(ps_gutMicro, variable = "EH_GL_C", title = "Entamoeba Histolytica, Giardia lamblia, and Cryptosporidium parvum Coinfection")
ANY <- a_diversity_categorical(ps_gutMicro, variable = "ANY", title = "Any Protozoan Infection")

```


## 4 - Beta Diversity (Bray & Jaccard)
```{r beta diversity}

# beta diversity pcoa plot w/ bray distance
# categorical variable
beta_diversity_bray_plot_categorical <- function(ps, variable, title_name=variable) {
  
  # Convert sample data to data frame
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  
  # Remove NAs
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  
  # Updated sample data after NA removal
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  
  # Count occurrences of each value in the variable
  value_counts <- table(sam_data_filtered[[variable]])
  print(value_counts)  # View group sizes
  
  # Identify values that appear at least 3 times (PCOA plot can't draw ellipse with < 3 points)
  valid_values <- names(value_counts)[value_counts >= 3]
  
  # Filter samples to only include those with valid values
  final_keep_samples <- sam_data_filtered[[variable]] %in% valid_values
  ps_final <- prune_samples(final_keep_samples, ps_filtered)
  
  # Get final sample data
  sam_data_final <- as(phyloseq::sample_data(ps_final), "data.frame")
  
  # Perform PCoA
  pcoares <- get_pcoa(obj = ps_final, distmethod = "bray", method = "hellinger")
  
  # Perform PERMANOVA to get p-value
  physeq_dist <- phyloseq::distance(ps_final, method = "bray")
  permanova <- vegan::adonis2(physeq_dist ~ sam_data_final[[variable]])
  p_value <- permanova$`Pr(>F)`[1]
  
  # Format p-value for display
  p_text <- ifelse(p_value < 0.001, "p < 0.001", 
                   paste("p =", format(round(p_value, 3), nsmall = 3)))
  
  # Create PCoA plot
  pcoaplot <- ggordpoint(obj = pcoares, biplot = FALSE, speciesannot = TRUE,
                         factorNames = c(variable), ellipse = TRUE, linesize = 1.5, 
                         ellipse_linewd = 1, ellipse_lty = 2) +
    ggtitle(title_name) + 
    guides(color=guide_legend(title=title_name, 
                              override.aes = list(size = 4))) +
    theme( 
          legend.title = element_blank(), 
          legend.text = element_text(size = 20)) +
    # Add p-value annotation
    annotate("text", x = Inf, y = Inf, 
             label = p_text, 
             hjust = 1.1, vjust = 1.5, 
             size = 4, fontface = "bold") +
    labs(color = paste("Kitchen Floor Material"))
  
  print(pcoaplot)
  return(pcoaplot)
}

beta_diversity_bray_plot_continuous <- function(ps, variable, title_name = variable) {
  # Convert sample data to data frame
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  
  # Remove NAs in the variable
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  
  # Check variable is numeric/continuous
  if(!is.numeric(sam_data_filtered[[variable]])) {
    stop("Selected variable must be numeric/continuous!")
  }
  
  # Perform PCoA
  ord <- ordinate(ps_filtered, method = "PCoA", distance = "bray")
  pco_df <- as.data.frame(ord$vectors)[, 1:2]
  pco_df$SampleID <- rownames(pco_df)
  pco_df$Variable <- sam_data_filtered[rownames(pco_df), variable]
  
  # Calculate PERMANOVA p-value (on continuous variable)
  physeq_dist <- phyloseq::distance(ps_filtered, method = "bray")
  ad <- vegan::adonis2(physeq_dist ~ sam_data_filtered[[variable]])
  p_value <- ad$`Pr(>F)`[1]
  p_text <- ifelse(p_value < 0.001, "p < 0.001", paste("p =", format(round(p_value, 3), nsmall = 3)))
  
  # Plot
  plt <- ggplot(pco_df, aes(x = Axis.1, y = Axis.2, color = Variable)) +
    geom_point(size = 4, alpha = 0.9) +
    scale_color_viridis_c(option = "plasma", name = variable) +
    labs(
      title = title_name,
      subtitle = p_text,
      x = "PCoA Axis 1",
      y = "PCoA Axis 2"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)
    )
  print(plt)
  return(plt)
}

beta_diversity_jaccard_plot_categorical <- function(ps, variable, title_name=variable) {
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  value_counts <- table(sam_data_filtered[[variable]])
  print(value_counts)
  valid_values <- names(value_counts)[value_counts >= 3]
  final_keep_samples <- sam_data_filtered[[variable]] %in% valid_values
  ps_final <- prune_samples(final_keep_samples, ps_filtered)
  sam_data_final <- as(phyloseq::sample_data(ps_final), "data.frame")
  
  # PCoA with Jaccard
  pcoares <- get_pcoa(obj = ps_final, distmethod = "jaccard", method = "hellinger")
  
  # PERMANOVA with Jaccard
  physeq_dist <- phyloseq::distance(ps_final, method = "jaccard")
  permanova <- vegan::adonis2(physeq_dist ~ sam_data_final[[variable]])
  p_value <- permanova$`Pr(>F)`[1]
  p_text <- ifelse(p_value < 0.001, "p < 0.001", 
                   paste("p =", format(round(p_value, 3), nsmall = 3)))
  
  pcoaplot <- ggordpoint(obj = pcoares, biplot = FALSE, speciesannot = TRUE,
                         factorNames = c(variable), ellipse = TRUE, linesize = 1.5, 
                         ellipse_linewd = 1, ellipse_lty = 2) +
    ggtitle(title_name) + 
    guides(color=guide_legend(title=title_name, override.aes = list(size = 4))) +
    theme(legend.title = element_blank(), legend.text = element_text(size = 20)) +
    annotate("text", x = Inf, y = Inf, 
             label = p_text, 
             hjust = 1.1, vjust = 1.5, 
             size = 4, fontface = "bold") +
    labs(color = variable)
  print(pcoaplot)
  return(pcoaplot)
}

beta_diversity_jaccard_plot_continuous <- function(ps, variable, title_name = variable) {
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  if(!is.numeric(sam_data_filtered[[variable]])) {
    stop("Selected variable must be numeric/continuous!")
  }
  
  # PCoA with Jaccard
  ord <- ordinate(ps_filtered, method = "PCoA", distance = "jaccard")
  pco_df <- as.data.frame(ord$vectors)[, 1:2]
  pco_df$SampleID <- rownames(pco_df)
  pco_df$Variable <- sam_data_filtered[rownames(pco_df), variable]
  
  # PERMANOVA with Jaccard
  physeq_dist <- phyloseq::distance(ps_filtered, method = "jaccard")
  ad <- vegan::adonis2(physeq_dist ~ sam_data_filtered[[variable]])
  p_value <- ad$`Pr(>F)`[1]
  p_text <- ifelse(p_value < 0.001, "p < 0.001", paste("p =", format(round(p_value, 3), nsmall = 3)))
  
  plt <- ggplot(pco_df, aes(x = Axis.1, y = Axis.2, color = Variable)) +
    geom_point(size = 4, alpha = 0.9) +
    scale_color_viridis_c(option = "plasma", name = variable) +
    labs(
      title = title_name,
      subtitle = p_text,
      x = "PCoA Axis 1",
      y = "PCoA Axis 2"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12)
    )
  print(plt)
  return(plt)
}

for (var in variables1){
  A <- beta_diversity_bray_plot_categorical(ps_gutMicro, var)
  B <- beta_diversity_jaccard_plot_categorical(ps_gutMicro, var)
}

EHbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "EH", title = "Entamoeba histolytica")
EHjacard <- beta_diversity_jaccard_plot_categorical(ps_gutMicro, variable = "EH", title = "Entamoeba histolytica")
GLbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "GL", title = "Giardia lamblia")
GLjacard <- beta_diversity_jaccard_plot_categorical(ps_gutMicro, variable = "GL", title = "Giardia lamblia")
Cbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "C", title = "Cryptosporidium")
Cjacard <- beta_diversity_jaccard_plot_categorical(ps_gutMicro, variable = "C", title = "Cryptosporidium")
ANYbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "ANY", title = " ")
ANYjacard <- beta_diversity_jaccard_plot_categorical(ps_gutMicro, variable = "ANY", title = " ")

ggsave("EH_beta.png", plot = EHbray, width = 6, height = 4, dpi = 500)
ggsave("GL_beta.png", plot = GLbray, width = 6, height = 4, dpi = 500)
ggsave("C_beta.png", plot = Cbray, width = 6, height = 4, dpi = 500)
ggsave("ANY_beta.png", plot = ANYbray, width = 6, height = 4, dpi = 500)



EH_GLbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "EH_GL", title = "EH_GL")
EH_Cbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "EH_C", title = "EH_C")
GL_Cbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "GL_C", title = "GL_C")
EH_GL_Cbray <- beta_diversity_bray_plot_categorical(ps_gutMicro, variable = "EH_GL_C", title = "EH_GL_C")

ggsave("EH_GL_beta.png", plot = EH_GLbray, width = 6, height = 4, dpi = 500)
ggsave("EH_C_beta.png", plot = EH_Cbray, width = 6, height = 4, dpi = 500)
ggsave("GL_C_beta.png", plot = GL_Cbray, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C_beta.png", plot = EH_GL_Cbray, width = 6, height = 4, dpi = 500)


```

## 5 - Relative Abundance Barplot (Genus & Family)
```{r relative abundance}

family_barplot_categorical <- function(ps, variable, title_name=variable) {
  
  # Convert sample data to data frame
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  
  # First remove NAs
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  
  # Get updated sample data after NA removal
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  
  # Count occurrences of each value in the variable
  value_counts <- table(sam_data_filtered[[variable]])
  
  # Identify values that appear at least 5 times
  valid_values <- names(value_counts)[value_counts >= 2]
  
  # Filter samples to only include those with valid values
  final_keep_samples <- sam_data_filtered[[variable]] %in% valid_values
  ps_final <- prune_samples(final_keep_samples, ps_filtered)
  
  familytaxa <- get_taxadf(obj = ps_final, taxlevel = 5)
  barplot <- ggbartax(obj = familytaxa, facetNames = variable, plotgroup = TRUE, topn = 10) +
    xlab(NULL) +
    ylab("Relative Abundance (%)") +
    labs(title = title_name) +  
    scale_fill_manual(values = c(colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(31))) +
    guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5, ncol = 2))
  
  
  print(barplot)
  return(barplot)
}

genus_barplot_categorical <- function(ps, variable, title_name=variable) {
  
  # Convert sample data to data frame
  sam_data <- as(phyloseq::sample_data(ps), "data.frame")
  
  # First remove NAs
  keep_samples <- !is.na(sam_data[[variable]])
  ps_filtered <- prune_samples(keep_samples, ps)
  
  # Get updated sample data after NA removal
  sam_data_filtered <- as(phyloseq::sample_data(ps_filtered), "data.frame")
  
  # Count occurrences of each value in the variable
  value_counts <- table(sam_data_filtered[[variable]])
  
  # Identify values that appear at least 5 times
  valid_values <- names(value_counts)[value_counts >= 2]
  
  # Filter samples to only include those with valid values
  final_keep_samples <- sam_data_filtered[[variable]] %in% valid_values
  ps_final <- prune_samples(final_keep_samples, ps_filtered)
  
  genustaxa <- get_taxadf(obj = ps_final, taxlevel = 6)
  barplot <- ggbartax(obj = genustaxa, facetNames = variable, plotgroup = TRUE, topn = 15) +
    xlab(NULL) +
    ylab("Relative Abundance (%)") +
    labs(title = title_name, fill = NULL) +  
    scale_fill_manual(values = c(colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(31))) +
    guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5, ncol = 2)) +
    theme_classic() +
    theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))
  
  print(barplot)
  return(barplot)
}

for (var in variables1){
  A <- genus_barplot_categorical(ben_phyloseq, var)
  B <- family_barplot_categorical(ben_phyloseq, var)
}


EHrelab <- genus_barplot_categorical(ps_gutMicro, variable = "EH", title = " ")
GLrelab <- genus_barplot_categorical(ps_gutMicro, variable = "GL", title = " ")
Crelab <- genus_barplot_categorical(ps_gutMicro, variable = "C", title = " ")
ANYrelab <- genus_barplot_categorical(ps_gutMicro, variable = "ANY", title = " ")

EH_GLrelab <- genus_barplot_categorical(ps_gutMicro, variable = "EH_GL", title = " ")
EH_Crelab <- genus_barplot_categorical(ps_gutMicro, variable = "EH_C", title = " ")
GL_Crelab <- genus_barplot_categorical(ps_gutMicro, variable = "GL_C", title = " ")
EH_GL_Crelab <- genus_barplot_categorical(ps_gutMicro, variable = "EH_GL_C", title = " ")

ggsave("EH_GL_relab.png", plot = EH_GLrelab, width = 6, height = 4, dpi = 500)
ggsave("EH_C_relab.png", plot = EH_Crelab, width = 6, height = 4, dpi = 500)
ggsave("GL_C_relab.png", plot = GL_Crelab, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C_relab.png", plot = EH_GL_Crelab, width = 6, height = 4, dpi = 500)

ggsave("EH_relab.png", plot = EHrelab, width = 6, height = 4, dpi = 500)
ggsave("GL_relab.png", plot = GLrelab, width = 6, height = 4, dpi = 500)
ggsave("C_relab.png", plot = Crelab, width = 6, height = 4, dpi = 500)
ggsave("ANY_relab.png", plot = ANYrelab, width = 6, height = 4, dpi = 500)


```

## Differential abundance volcano plot
```{r maaslin2}
create_volcano_plot <- function(ps_obj, taxlevel, condition_col, title_name = condition_col) {
  # Extract sample data, OTU table, taxonomy
  metadata <- as.data.frame(sample_data(ps_obj))
  SVs <- as.data.frame(otu_table(ps_obj))
  taxonomy <- as.data.frame(phyloseq::tax_table(ps_obj))
  
  # Accept taxlevel as number or name
  if (is.numeric(taxlevel)) {
    tax_col <- colnames(taxonomy)[taxlevel]
  } else {
    tax_col <- taxlevel
  }
  
  # Prepare grouping variable
  var_values <- metadata[[condition_col]]
  non_na_values <- var_values[!is.na(var_values)]
  value_counts <- table(non_na_values)
  
  # Top 2 most frequent (works for binary factors too)
  top_values <- names(sort(value_counts, decreasing = TRUE))[1:2]
  subset_idx <- var_values %in% top_values & !is.na(var_values)
  metadata <- metadata[subset_idx, , drop = FALSE]
  
  # Subset SVs to match samples
  sample_names_to_use <- rownames(metadata)
  # Match sample orientation
  if (taxa_are_rows(ps_obj)) {
    SVs <- SVs[, sample_names_to_use, drop = FALSE]
  } else {
    SVs <- SVs[sample_names_to_use, , drop = FALSE]
    SVs <- t(SVs)
  }
  
  # Ensure that taxonomy contains the specified level
  taxonomy <- taxonomy %>%
    tibble::rownames_to_column(var = "OTU_name") %>%
    dplyr::select(OTU_name, !!sym(tax_col)) %>%
    distinct()
  
  # Combine SVs with taxonomy
  SVs_with_taxonomy <- SVs %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "OTU_name") %>%
    left_join(taxonomy, by = "OTU_name")
  
  # Aggregate by taxonomic level
  tax_level_SVs <- SVs_with_taxonomy %>%
    dplyr::select(-OTU_name) %>%
    group_by(!!sym(tax_col)) %>%
    summarise(across(everything(), ~sum(.x, na.rm = TRUE))) %>%
    filter(!is.na(!!sym(tax_col)), !!sym(tax_col) != "none") %>%
    tibble::column_to_rownames(var = tax_col)
  
  # Ensure column/sample order matches metadata
  tax_level_SVs <- tax_level_SVs[, rownames(metadata)]
  
  # Convert to matrix for ALDEx2
  tax_level_SVs <- as.matrix(tax_level_SVs)
  condition <- as.character(metadata[[condition_col]])
  
  # Run ALDEx2 analysis
  aldex_data <- aldex(tax_level_SVs, conditions = condition, mc.samples = 128, test = "t", effect = TRUE)
  results <- data.frame(aldex_data)
  results <- results %>% tibble::rownames_to_column(var = tax_col)
  results <- results %>% arrange(we.eBH)
  
  # Volcano plot
  p <- results %>%
    mutate(Significant = we.eBH < 0.05) %>%
    mutate(Taxon = as.character(!!sym(tax_col))) %>%
    mutate(TaxonToPrint = if_else(Significant, paste0(Taxon, " (", round(we.eBH, 3), ")"), "")) %>%
    ggplot(aes(x = diff.btw, y = -log10(we.ep), color = Significant, label = TaxonToPrint)) +
    geom_point(alpha = 0.6, shape = 16) +
    geom_text_repel(size = 4, nudge_y = 0.05, max.overlaps = Inf) +
    theme_minimal() +
    xlab("log2(fold change)") +
    ylab("-log10(P-value)") +
    scale_color_manual(values = c("black", "red")) +
    theme(legend.position = "none") +
    ggtitle(paste(title_name, "\nTaxonomic Level:", tax_col))
  
  print(p)
  return(list(plot = p, table = results))
}

plot_aldex2_boxplots <- function(
  ps_obj,
  aldex2_table,
  variable = "Condition",
  variable_name = variable,
  taxlevel = "Genus"
) {
  sig_genera <- aldex2_table %>%
    filter(we.eBH < 0.05) %>%
    pull(!!sym(taxlevel)) %>%
    unique()

  plot_single <- function(genus) {
    qval <- aldex2_table %>%
      filter(!!sym(taxlevel) == genus) %>%
      arrange(we.eBH) %>%
      slice(1) %>%
      pull(we.eBH)
    qval_label <- if (length(qval) == 0) "NA" else formatC(qval, digits = 3, format = "f")
    auto_title <- paste0(genus, " (q = ", qval_label, ")")

    otumat <- as.data.frame(otu_table(ps_obj))
    if (!taxa_are_rows(ps_obj)) otumat <- t(otumat)
    taxmat <- as.data.frame(phyloseq::tax_table(ps_obj))
    otumat$Taxon <- taxmat[rownames(otumat), taxlevel]
    
    feature_table <- otumat %>%
      group_by(Taxon) %>%
      summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
      tibble::column_to_rownames("Taxon")

    feature_table_rel <- sweep(feature_table, 2, colSums(feature_table), "/")

    if (!genus %in% rownames(feature_table_rel)) return(NULL)

    genus_abund <- as.numeric(feature_table_rel[genus, ])
    sample_names <- colnames(feature_table_rel)
    meta <- as.data.frame(sample_data(ps_obj))[sample_names, , drop = FALSE]

    df <- data.frame(
      Sample = sample_names,
      Abundance = genus_abund,
      Group = as.factor(meta[[variable]])
    )

    ggplot(df, aes(x = Group, y = Abundance, fill = Group)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6, color = "black") +
      labs(
        title = auto_title,
        x = variable_name,
        y = "Relative Abundance"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
  }

  plot_list <- lapply(sig_genera, plot_single)
  names(plot_list) <- sig_genera
  plot_list <- plot_list[!sapply(plot_list, is.null)]
  message(sprintf("Created %d boxplots for significant genera.", length(plot_list)))
  return(plot_list)
}


run_maaslin2_and_plot <- function(
  ps_obj, 
  variable, 
  title_name = variable, 
  taxlevel = "Genus",
  qval_sig_max = 0.1,
  min_prevalence = 0.05,
  min_samples = 5,               # Minimum number of samples with abundance > 0
  min_abundance_mean = 0.0005,    # Minimum mean relative abundance across all samples
  min_abundance_max  = 0.001,     # Minimum *max* relative abundance in any sample
  adjust_vars = c("Age", "sex"),
  remove_unassigned = TRUE
) {
  
 

  # Prepare feature table (taxa as rows, samples as columns)
  otumat <- as.data.frame(otu_table(ps_obj))
  if (!taxa_are_rows(ps_obj)) otumat <- t(otumat)
  
  # Aggregate at desired taxonomic level
  taxmat <- as.data.frame(phyloseq::tax_table(ps_obj))
  if (!(taxlevel %in% colnames(taxmat))) stop("taxlevel not in taxonomy table!")
  
  # Assign "Unassigned" to missing genus
  otumat$Taxon <- taxmat[rownames(otumat), taxlevel]
  otumat$Taxon[is.na(otumat$Taxon) | otumat$Taxon == "" ] <- "Unassigned"
  
  # Aggregate counts at genus level
  feature_table <- otumat %>%
    group_by(Taxon) %>%
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
    tibble::column_to_rownames("Taxon")
  
  # Convert to relative abundance (to apply abundance filters)
  feature_table_rel <- sweep(feature_table, 2, colSums(feature_table), "/")
  
  # Filtering step: prevalence, mean abundance, and max abundance
  present_counts <- rowSums(feature_table_rel > 0)
  mean_abund <- rowMeans(feature_table_rel)
  max_abund <- apply(feature_table_rel, 1, max)
  
  keep_taxa <- (present_counts > (min_prevalence * ncol(feature_table_rel))) &
               (present_counts >= min_samples) &
               (mean_abund >= min_abundance_mean | max_abund >= min_abundance_max)
  
  feature_table <- feature_table[keep_taxa, ]
  feature_table_rel <- feature_table_rel[keep_taxa, ]
  
  # Prepare metadata
  meta <- as(phyloseq::sample_data(ps_obj), "data.frame")
  meta <- meta[ , , drop = FALSE]
  
  # Match samples between metadata and feature table
  common_samples <- intersect(colnames(feature_table), rownames(meta))
  feature_table <- feature_table[ , common_samples, drop = FALSE]
  meta <- meta[common_samples, , drop = FALSE]
  
  # Compose fixed effects
  meta$Age <- as.numeric(as.character(meta$Age))
  fixed_effects <- unique(c(variable, adjust_vars))
  fixed_effects <- fixed_effects[fixed_effects %in% colnames(meta)]
  
  # Run MaAsLin2 
  fit_data <- Maaslin2(
    input_data = as.data.frame(t(feature_table)),
    input_metadata = meta,
    output = tempdir(),   
    fixed_effects = fixed_effects,
    normalization = "TSS",
    min_prevalence = min_prevalence
  )
  
  # Results
  results <- fit_data$results
  results_var <- results[results$metadata == variable, ]
  results_var <- results_var %>%
    dplyr::mutate(sig = ifelse(qval < qval_sig_max, "Significant", "NS"))
  
  # Only keep one result per genus
  results_var_unique <- results_var %>%
    group_by(feature) %>%
    slice_min(order_by = qval, n = 1, with_ties = FALSE) %>%
    ungroup()
  
  # Prevalence and abundance calculation (add to output)
  prevalence_df <- data.frame(
    feature = rownames(feature_table_rel),
    prevalence = rowSums(feature_table_rel > 0),
    prevalence_frac = rowSums(feature_table_rel > 0) / ncol(feature_table_rel),
    mean_abund = rowMeans(feature_table_rel),
    max_abund = apply(feature_table_rel, 1, max)
  )
  results_var_unique <- left_join(results_var_unique, prevalence_df, by = "feature")
  
  # Keep Only Genus Names That are in the Original Genus Column
  genus_list <- unique(taxmat[[taxlevel]])
  genus_list <- genus_list[!is.na(genus_list) & genus_list != ""]
  if (remove_unassigned) {
    genus_list <- genus_list[genus_list != "Unassigned"]
  }
  results_var_unique <- results_var_unique %>% filter(feature %in% genus_list)
  
  # Plot: Label rare significant genera
  results_var_unique <- results_var_unique %>%
    mutate(
      label_flag = ifelse(
        sig == "Significant" & prevalence < min_samples, 
        paste0(feature, " (rare)"), 
        ifelse(sig == "Significant", feature, "")
      )
    )
  
  p <- ggplot(results_var_unique, aes(x = coef, y = -log10(qval), color = sig)) +
    geom_point() +
    geom_text_repel(
      aes(label = label_flag),
      size = 3, 
      max.overlaps = 12, 
      color = "red",
      force = 2,
      box.padding = 0.5
    ) +
    scale_color_manual(values = c("Significant" = "red", "NS" = "black")) +
    labs(
      x = "Effect Size Coefficient",
      y = "-log10(FDR)",
      title = title_name
    ) +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(p)
  n_rare_sig <- sum(results_var_unique$sig == "Significant" & results_var_unique$prevalence < min_samples)
  if (n_rare_sig > 0) {
    message(sprintf("Warning: %d significant genera are present in fewer than %d samples (labeled as (rare)). Interpret with caution!", n_rare_sig, min_samples))
  }
  return(list(plot = p, table = results_var_unique, all_results = results))
}

plot_all_significant_boxplots <- function(
  ps_obj, 
  maaslin2_table, 
  variable = "FoodSecure_vs_FoodInsecure", 
  variable_name = variable,
  taxlevel = "Genus"
) {
  
  sig_genera <- maaslin2_table %>%
    filter(sig == "Significant", metadata == variable) %>%
    pull(feature) %>%
    unique()
  
  # Helper function for a single genus
  plot_single <- function(genus) {
    qval <- maaslin2_table %>%
      filter(feature == genus, metadata == variable) %>%
      arrange(qval) %>%
      slice(1) %>%
      pull(qval)
    qval_label <- if(length(qval) == 0) "NA" else formatC(qval, digits = 3, format = "f")
    auto_title <- paste0(genus, " (q = ", qval_label, ")")
    
    otumat <- as.data.frame(otu_table(ps_obj))
    if (!taxa_are_rows(ps_obj)) otumat <- t(otumat)
    taxmat <- as.data.frame(phyloseq::tax_table(ps_obj))
    otumat$Taxon <- taxmat[rownames(otumat), taxlevel]
    feature_table <- otumat %>%
      group_by(Taxon) %>%
      summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
      tibble::column_to_rownames("Taxon")
    feature_table_rel <- sweep(feature_table, 2, colSums(feature_table), "/")
    if (!genus %in% rownames(feature_table_rel)) return(NULL)
    genus_abund <- as.numeric(feature_table_rel[genus, ])
    sample_names <- colnames(feature_table_rel)
    meta <- as(phyloseq::sample_data(ps_obj), "data.frame")
    meta <- meta[sample_names, , drop = FALSE]
    df <- data.frame(
      Sample = sample_names,
      Abundance = genus_abund,
      Group = as.factor(meta[[variable]])
    )
    ggplot(df, aes(x = Group, y = Abundance, fill = Group)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6, color = "black") +
      labs(
        title = auto_title,
        x = variable_name,
        y = "Relative Abundance"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
  }
  
  # Loop and collect
  plot_list <- lapply(sig_genera, plot_single)
  names(plot_list) <- sig_genera
  plot_list <- plot_list[!sapply(plot_list, is.null)]  # Remove any nulls
  print(paste("Created", length(plot_list), "boxplots for significant genera."))
  plot_list
}

#for (var in variables1){
  #var_maaslin2 <- run_maaslin2_and_plot(ps_gutMicro, var, qval_sig_max = 0.2) 
  #var_boxplots <- plot_all_significant_boxplots(
    #ps_obj = ps_gutMicro,
    #maaslin2_table = var_maaslin2$table,
    #variable = var)
  #print(var_boxplots[[1]])
#}

for (var in variables1){
  create_volcano_plot(ps_gutMicro, "Genus", var)
  #print(var_boxplots[[1]])
}

#Run ALDEx2 plot and get the result table
volcano_out <- create_volcano_plot(ps_gutMicro, taxlevel = "Genus", condition_col = "EH_GL_C")
aldex2_table <- volcano_out$table

 #Generate boxplots for significant genera
aldex_boxplots <- plot_aldex2_boxplots(
  ps_obj = ps_gutMicro,
  aldex2_table = aldex2_table,
  variable = "EH_GL_C"
)

# View one plot
print(aldex_boxplots[[2]])

ggsave("EH_GL1.png", plot = aldex_boxplots[[1]], width = 6, height = 4, dpi = 500)


ggsave("EH_GL_C1.png", plot = aldex_boxplots[[1]], width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C2.png", plot = aldex_boxplots[[2]], width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C3.png", plot = aldex_boxplots[[3]], width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C4.png", plot = aldex_boxplots[[4]], width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C5.png", plot = aldex_boxplots[[5]], width = 6, height = 4, dpi = 500)


ggsave("EH_GL_C_volcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("EH_GLvolcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("EH_volcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("GL_volcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("C_volcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("EH_Cvolcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("GL_Cvolcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("ANYvolcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_Cvolcano.png", plot = volcano_out$plot, width = 6, height = 4, dpi = 500)
```

## ML Model Training 
```{r ML training}
cv_predict_clr_xgb <- function(
  ps_obj,
  outcome_var,
  meta_cols = NULL,
  min_prevalence = 0.05,
  nfolds = 10,
  seed = 100,
  n_cores = 4
) {
  library(parallel)
  library(doParallel)
  library(phyloseq)

  set.seed(seed)
  cl <- parallel::makeCluster(n_cores)  # <- fix here
  registerDoParallel(cl)

  get_tax_matrix <- function(ps, taxlevel, prefix) {
    taxlevel <- as.character(taxlevel)  # <- ensures it's not a list
  
    ps_tax <- tax_glom(ps, taxlevel)
    otumat <- as.data.frame(otu_table(ps_tax))
    if (!taxa_are_rows(ps_tax)) otumat <- t(otumat)
  
    taxmat <- as.data.frame(phyloseq::tax_table(ps_tax))
  
    if (!(taxlevel %in% colnames(taxmat))) {
      stop(paste("Taxlevel", taxlevel, "not found in tax_table columns."))
    }
  
    tax_names <- as.character(taxmat[[taxlevel]])
    tax_names[is.na(tax_names) | tax_names == ""] <- "Unassigned"
    rownames(otumat) <- paste0(prefix, make.unique(tax_names))
  
    present_counts <- rowSums(otumat > 0)
    keep <- present_counts >= (min_prevalence * ncol(otumat))
    otumat <- otumat[keep, , drop = FALSE]
  
    return(otumat)
  }


  fam_tab <- get_tax_matrix(ps_obj, "Family", "F__")
  gen_tab <- get_tax_matrix(ps_obj, "Genus", "G__")
  all_feat_tab <- rbind(fam_tab, gen_tab)
  all_feat_tab[all_feat_tab == 0] <- 1e-6
  clr_mat <- compositions::clr(all_feat_tab)
  clr_mat <- t(clr_mat)

  meta <- as.data.frame(phyloseq::sample_data(ps_obj))
  clr_samples <- rownames(clr_mat)
  meta <- meta[clr_samples, , drop = FALSE]
  y <- as.factor(meta[[outcome_var]])
  keep <- !is.na(y)
  clr_mat <- clr_mat[keep, , drop = FALSE]
  meta <- meta[keep, , drop = FALSE]
  y <- y[keep]
  levels(y) <- make.names(levels(y))

  if (length(meta_cols) > 0 && any(meta_cols %in% colnames(meta))) {
    meta_cols <- intersect(meta_cols, colnames(meta))
    meta_features <- meta[, meta_cols, drop = FALSE]
    meta_features[] <- lapply(meta_features, function(x) as.numeric(as.character(x)))
    meta_features <- meta_features[, sapply(meta_features, function(x) all(x %in% c(0, 1, NA))), drop = FALSE]
    X_full <- cbind(clr_mat, meta_features)
  } else {
    X_full <- clr_mat
    meta_cols <- character(0)  # ensure later steps don't break
  }

  
  nzv <- caret::nearZeroVar(X_full)
  if (length(nzv) > 0) X_full <- X_full[, -nzv, drop = FALSE]

  # hyperparameter grid
  xgb_grid <- expand.grid(
    nrounds = c(100, 200),
    max_depth = c(3, 5, 7),
    eta = c(0.05, 0.1),
    gamma = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample = 0.8
  )

  fitControl <- caret::trainControl(
    method = "cv",
    number = nfolds,
    classProbs = TRUE,
    savePredictions = "final",
    allowParallel = TRUE
  )

  xgb_fit <- caret::train(
    x = X_full,
    y = y,
    method = "xgbTree",
    tuneGrid = xgb_grid,
    trControl = fitControl,
    verbose = FALSE
  )

  overall_acc <- max(xgb_fit$results$Accuracy)
  best_params <- xgb_fit$bestTune

  varimp <- caret::varImp(xgb_fit, scale = FALSE)$importance
  varimp_df <- tibble(Feature = rownames(varimp), Importance = varimp[,1]) %>%
    arrange(desc(Importance)) %>%
    mutate(Level = dplyr::case_when(
      grepl("^F__", Feature) ~ "Family",
      grepl("^G__", Feature) ~ "Genus",
      Feature %in% meta_cols ~ "Metadata",
      TRUE ~ "Other"
    ))

  preds <- xgb_fit$pred$pred[order(xgb_fit$pred$rowIndex)]
  y_true <- xgb_fit$pred$obs[order(xgb_fit$pred$rowIndex)]
  overall_cm <- caret::confusionMatrix(preds, y_true)

  stopCluster(cl)
  registerDoSEQ()

  return(list(
    overall_accuracy = overall_acc,
    best_params = best_params,
    confusion_matrix = overall_cm,
    feature_importance = varimp_df,
    predictions = preds,
    y_true = y_true,
    xgb_fit = xgb_fit
  ))
}

# E.g.
results_xgb <- cv_predict_clr_xgb(ps_gutMicro, "GL")
results_xgb$confusion_matrix
head(results_xgb$feature_importance, 50)

```

## Linear Discriminant analysis w/ Lefse
```{r diff Lefse}
create_lefse_barplot <- function(ps, factor, title_name = factor, lda_cut = 3, wil_cut = 0.01, kw_cut = 0.01) {
  # Extract sample data
  sample_df <- data.frame(sample_data(ps))

  # Get the variable of interest
  var_vector <- sample_df[[factor]]

  # Exclude NAs
  var_vector_no_na <- var_vector[!is.na(var_vector)]

  keep_samples <- rownames(sample_df)[!is.na(var_vector)]
  ps_subset <- prune_samples(keep_samples, ps)
  
  # Run LEfSe
  mm_lefse <- tryCatch(
    run_lefse(
      ps_subset,
      group = factor,
      wilcoxon_cutoff = wil_cut,
      kw_cutoff = kw_cut,
      lda_cutoff = lda_cut,
      multigrp_strat = TRUE
    ),
    error = function(e) {
      message(paste("LefSe failed for", factor, ":", e$message))
      return(NULL)
    }
  )

  # Remove species-level features (those containing 's__')
  if (!is.null(mm_lefse)) {
    mm_lefse@marker_table <- mm_lefse@marker_table[!grepl("s__", mm_lefse@marker_table$feature), ]
  }

  # Plot or fallback
  if (!is.null(mm_lefse) && nrow(mm_lefse@marker_table) > 0) {
    lefse_bar_plot <- tryCatch(
      plot_ef_bar(mm_lefse) +
        ggtitle(title_name),
      error = function(e) {
        message(paste("Failed to create LefSe bar plot:", e$message))
        ggplot() + annotate("text", x = 1, y = 1, label = "LefSe plot failed", size = 6) + theme_void()
      }
    )
  } else {
    lefse_bar_plot <- ggplot() +
      annotate("text", x = 1, y = 1, label = "No significant non-species features", size = 6) +
      theme_void()
  }

  print(lefse_bar_plot)
  return(lefse_bar_plot)
}


EH_lefse<- create_lefse_barplot(ps_gutMicro, "EH", "EH", lda_cut = 2, wil_cut = 0.05, kw_cut = 0.05)
GL_lefse<- create_lefse_barplot(ps_gutMicro, "GL", "GL", lda_cut = 2, wil_cut = 0.05, kw_cut = 0.05)
C_lefse<- create_lefse_barplot(ps_gutMicro, "C", "C", lda_cut = 2, wil_cut = 0.03, kw_cut = 0.03)
ANY_lefse<- create_lefse_barplot(ps_gutMicro, "ANY", "ANY", lda_cut = 2, wil_cut = 0.05, kw_cut = 0.05)
EH_GL_lefse<- create_lefse_barplot(ps_gutMicro, "EH_GL", "EH_GL", lda_cut = 2, wil_cut = 0.05, kw_cut = 0.05)
EH_C_lefse<- create_lefse_barplot(ps_gutMicro, "EH_C", "EH_C", lda_cut = 2, wil_cut = 0.1, kw_cut = 0.1)
GL_C_lefse<- create_lefse_barplot(ps_gutMicro, "GL_C", "GL_C", lda_cut = 4, wil_cut = 0.05, kw_cut = 0.05)
EH_GL_C_lefse<- create_lefse_barplot(ps_gutMicro, "EH_GL_C", "EH_GL_C", lda_cut = 2, wil_cut = 0.05, kw_cut = 0.05)



ggsave("EH_lefse.png", plot = EH_lefse, width = 6, height = 4, dpi = 500)
ggsave("GL_lefse.png", plot = GL_lefse, width = 6, height = 4, dpi = 500)
ggsave("C_lefse.png", plot = C_lefse, width = 6, height = 4, dpi = 500)
ggsave("ANY_lefse.png", plot = ANY_lefse, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_lefse.png", plot = EH_GL_lefse, width = 6, height = 4, dpi = 500)
ggsave("EH_C_lefse.png", plot = EH_C_lefse, width = 6, height = 4, dpi = 500)
ggsave("GL_C_lefse.png", plot = GL_C_lefse, width = 6, height = 4, dpi = 500)
ggsave("EH_GL_C_lefse.png", plot = EH_GL_C_lefse, width = 6, height = 4, dpi = 500)


```


## ROC, Variable Importance, and Heatmap plots
```{r ml plots}

plot_roc_curve_gg <- function(model_results, positive_class = NULL) {
  
  y_true <- model_results$y_true
  if (is.null(positive_class)) {
    positive_class <- levels(y_true)[1]
  }
  pred_prob <- model_results$xgb_fit$pred %>%
    filter(obs %in% levels(y_true)) %>%
    arrange(rowIndex) %>%
    pull(!!as.name(positive_class))
  
  roc_obj <- roc(y_true, pred_prob, levels = rev(levels(y_true)), direction = "<")
  auc_value <- auc(roc_obj)
  
  ggroc(roc_obj, legacy.axes = TRUE, colour = "darkgreen", size = 1.3) +
    geom_abline(linetype = "dashed", color = "gray50") +
    labs(
      title = sprintf("ROC Curve (AUC = %.2f)", auc_value),
      x = "False Positive Rate (1 - Specificity)",
      y = "True Positive Rate (Sensitivity)"
    ) +
    theme_minimal(base_size = 14)
}

plot_top_importance <- function(model_results, n_top = 10, bar_color = "#2c7bb6") {
  
  varimp <- model_results$feature_importance %>% 
    arrange(desc(Importance)) %>%
    head(n_top)
  
  ggplot(varimp, aes(x = reorder(Feature, Importance), y = Importance)) +
    geom_bar(stat = "identity", fill = bar_color) +
    coord_flip() +
    labs(
      title = paste("Top", n_top, "Feature Importance"),
      x = "",
      y = "Importance"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
}

plot_top_feature_heatmap_clr <- function(
  ps_obj,
  model_results,
  n_top = 10,
  metadata_vars = NULL,
  outcome_var = "FoodSecure_vs_FoodInsecure",
  min_prevalence = 0.05
) {
  
  # Top N features
  top_features <- head(model_results$feature_importance$Feature, n_top)
  
  # Genus & family tables
  get_tax_matrix <- function(ps, taxlevel, prefix) {
    ps_tax <- tax_glom(ps, taxlevel)
    otumat <- as.data.frame(otu_table(ps_tax))
    if (!taxa_are_rows(ps_tax)) otumat <- t(otumat)
    taxmat <- as.data.frame(phyloseq::tax_table(ps_tax))
    tax_names <- as.character(taxmat[[taxlevel]])
    tax_names[is.na(tax_names) | tax_names == ""] <- "Unassigned"
    rownames(otumat) <- paste0(prefix, make.unique(tax_names))
    present_counts <- rowSums(otumat > 0)
    keep <- present_counts >= (min_prevalence * ncol(otumat))
    otumat <- otumat[keep, , drop=FALSE]
    return(otumat)
  }
  
  fam_tab <- get_tax_matrix(ps_obj, "Family", "F__")
  gen_tab <- get_tax_matrix(ps_obj, "Genus", "G__")
  all_feat_tab <- rbind(fam_tab, gen_tab)
  
  # CLR transformation
  all_feat_tab[all_feat_tab == 0] <- 1e-6
  clr_mat <- compositions::clr(all_feat_tab)
  clr_mat <- t(clr_mat)  # samples as rows, features as columns
  
  # Metadata (as numeric)
  meta <- as.data.frame(phyloseq::sample_data(ps_obj))
  clr_samples <- rownames(clr_mat)
  meta <- meta[clr_samples, , drop = FALSE]

  # Handling NaN
  if (!is.null(metadata_vars) && length(metadata_vars) > 0) {
    metadata_vars <- intersect(metadata_vars, colnames(meta))
    metadata_features <- meta[, metadata_vars, drop = FALSE]
    metadata_features[] <- lapply(metadata_features, function(x) as.numeric(as.character(x)))
    X_full <- cbind(clr_mat, metadata_features)
  } else {
    X_full <- clr_mat  # Just CLR matrix, no metadata
  }

  # Ensure all columns are numeric
  for (i in seq_len(ncol(X_full))) {
    if (!is.numeric(X_full[, i])) {
      X_full[, i] <- as.numeric(as.character(X_full[, i]))
    }
  }
  
  # Subset to top features, as rows (order preserved)
  top_feats_present <- top_features[top_features %in% colnames(X_full)]
  if (length(top_feats_present) == 0) stop("No top features found in input matrix.")
  heatmap_mat <- t(X_full[, top_feats_present, drop = FALSE])
  
  # Remove rows (features) that are all NA or all zero
  keep_rows <- apply(heatmap_mat, 1, function(x) any(!is.na(x)) && any(x != 0))
  heatmap_mat <- heatmap_mat[keep_rows, , drop = FALSE]
  if (nrow(heatmap_mat) == 0) stop("No valid features to plot after removing all-NA/zero rows.")
  
  # Annotation for sample outcome
  sample_anno <- as(phyloseq::sample_data(ps_obj), "data.frame")[, outcome_var, drop = FALSE]
  sample_anno <- sample_anno[colnames(heatmap_mat), , drop = FALSE]
  sample_anno[[outcome_var]] <- as.factor(as.character(sample_anno[[outcome_var]]))
  rownames(sample_anno) <- colnames(heatmap_mat)
  
  # Order columns by group
  group_order <- order(sample_anno[[outcome_var]])
  heatmap_mat <- heatmap_mat[, group_order, drop = FALSE]
  sample_anno <- sample_anno[group_order, , drop = FALSE]
  
  # Z-score (standardize) each feature (row)
  heatmap_mat_scaled <- t(scale(t(heatmap_mat)))
  heatmap_mat_scaled[is.na(heatmap_mat_scaled)] <- 0
  
  # Plot heatmap (no clustering of samples)
  png("top_features_heatmap.png", width = 5000, height = 1800, res = 500)
  pheatmap::pheatmap(
    mat = heatmap_mat_scaled,
    annotation_col = sample_anno,
    main = paste("Top", n_top, "Features Heatmap"),
    clustering_method = "complete",
    cluster_cols = FALSE,
    cluster_rows = TRUE,
    fontsize_row = 10,
    fontsize_col = 7,
    scale = "none",
    color = colorRampPalette(c("navy", "white", "firebrick3"))(100)
  )
  dev.off()
}


# E.g.
plot_top_feature_heatmap_clr(
  ps_obj = ps_gutMicro,
  model_results = results_xgb,
  n_top = 10,
  metadata_vars = NULL,
  outcome_var = "GL",
  min_prevalence = 0.05
)




library(pROC)
library(ggplot2)
plot_roc_curve_gg(results_xgb)
plot_top_importance(results_xgb, n_top = 10) 


roc_plot <- plot_roc_curve_gg(results_xgb)
ggsave("roc_curve.png", plot = roc_plot, width = 6, height = 4, dpi = 500)

# Generate feature importance plot
imp_plot <- plot_top_importance(results_xgb, n_top = 10)

# Save feature importance plot
ggsave("feature_importance.png", plot = imp_plot, width = 6, height = 5, dpi = 300)

```

## Picrust differential pathway analysis (NEED TO UPDATE)
```{r}
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tibble); library(stringr)
  library(phyloseq)
  library(ggplot2)
  library(ggprism)
  library(GGally)
  library(ggpicrust2)
})

# Helper: clean sample IDs
.clean_ids <- function(x){
  x <- as.character(x)
  x <- gsub("^UniqueID", "", x, ignore.case = TRUE)
  x <- gsub("^X(?=\\d)", "", x, perl = TRUE)
  trimws(x)
}

run_picrust_daa_metacyc <- function(
  file_name,
  ps_obj,
  group_col,
  recode_map   = NULL,
  p_adj_cutoff = 0.10,
  p_raw_cutoff = 0.05,
  top_n        = 20,
  min_nonzero  = 3
){
  # 1) Load & tidy abundance (MetaCyc)
  raw <- readr::read_tsv(file_name, show_col_types = FALSE)
  if (!"pathway" %in% names(raw))
    stop("The input table must contain a 'pathway' column (MetaCyc IDs).")

  abundance <- raw |>
    select(-any_of("description")) |>
    column_to_rownames("pathway") |>
    as.data.frame()

  # Clean sample IDs
  colnames(abundance) <- .clean_ids(colnames(abundance))

  # Ensure numeric
  suppressWarnings(abundance[] <- lapply(abundance, as.numeric))

  # Collapse duplicate sample IDs if any
  if (any(duplicated(colnames(abundance)))) {
    warning("Duplicated sample IDs after cleaning; collapsing by mean.")
    abund_t <- t(abundance)
    grp <- colnames(abundance)
    abund_sum <- rowsum(abund_t, group = grp)
    abundance <- t(abund_sum / as.vector(table(grp)))
  }

  # 2) Metadata
  meta <- as(sample_data(ps_obj), "data.frame") |>
    rownames_to_column("sample_name")

  if (!group_col %in% names(meta)) {
    stop(sprintf("Column '%s' not found in sample_data(ps_obj).", group_col))
  }

  if (!is.null(recode_map)) {
    meta[[group_col]] <- dplyr::recode(as.character(meta[[group_col]]), !!!recode_map)
  }
  meta[[group_col]] <- as.factor(meta[[group_col]])

  # 3) Align samples
  common <- intersect(colnames(abundance), meta$sample_name)
  message(sprintf("PICRUSt2 samples: %d | phyloseq samples: %d | overlap: %d",
                  ncol(abundance), nrow(meta), length(common)))
  if (length(common) < 3) {
    stop("Fewer than 3 overlapping samples after cleaning.")
  }

  abundance <- abundance[, common, drop = FALSE]
  meta      <- meta |> filter(sample_name %in% common)
  abundance <- abundance[, meta$sample_name, drop = FALSE]

  # Optional: prefilter sparse features
  if (min_nonzero > 1L) {
    keep <- rowSums(abundance > 0, na.rm = TRUE) >= min_nonzero
    abundance <- abundance[keep, , drop = FALSE]
  }

  # 4) DAA (LinDA on MetaCyc)
  daa_res <- pathway_daa(
    abundance  = abundance,
    metadata   = meta,
    group      = group_col,
    daa_method = "LinDA"
  )

  annot <- pathway_annotation(
    pathway        = "MetaCyc",
    daa_results_df = daa_res,
    ko_to_kegg     = FALSE
  )

  # 5) Detect p-value columns
  p_adj_col <- intersect(c("p_adjust","p.adjust","padj","q_value","qvalue"), names(annot))[1]
  p_col     <- intersect(c("p_values","p.value","pvalue","p"), names(annot))[1]
  if (is.na(p_adj_col)) stop("Adjusted p-value column not found in DAA results.")
  if (is.na(p_col))     p_col <- p_adj_col

  # Clean p-value table
  pval_table <- annot |>
    dplyr::transmute(
      feature,
      description = dplyr::coalesce(.data$description, feature),
      p_value     = .data[[p_col]],
      p_adjust    = .data[[p_adj_col]]
    )

  # 6) Check if adjusted p's are significant
  # --- choose which p-values to drive the plot threshold (fallback if none sig)
  n_sig_adj <- sum(annot[[p_adj_col]] <= p_adj_cutoff, na.rm = TRUE)
  use_raw_p_for_plot <- (is.na(n_sig_adj) || n_sig_adj == 0)

  if (use_raw_p_for_plot) {
    message(sprintf(
      "No pathways significant at adjusted p ≤ %.3f; using raw p-values for plotting.",
      p_adj_cutoff
    ))
    # Keep all columns, but for plotting, treat raw p as the adjusted p
    daa_for_plot <- annot
    daa_for_plot[[p_adj_col]] <- annot[[p_col]]   # <- key line

    # Rank/top-N by raw p-values
    top_feats <- annot |>
      dplyr::arrange(.data[[p_col]]) |>
      dplyr::slice_head(n = top_n) |>
      dplyr::pull(feature)

    plot_threshold <- p_raw_cutoff  # use your raw-p cutoff here
  } else {
    daa_for_plot <- annot
    # Rank/top-N by adjusted p (tie-break by raw p)
    top_feats <- annot |>
      dplyr::arrange(.data[[p_adj_col]], .data[[p_col]]) |>
      dplyr::slice_head(n = top_n) |>
      dplyr::pull(feature)

    plot_threshold <- p_adj_cutoff
  }


  # 7) Plot
  p_err <- pathway_errorbar(
    abundance          = abundance,
    daa_results_df     = daa_for_plot,
    Group              = meta[[group_col]],
    p_values_threshold = plot_threshold,
    order              = "group",
    select             = top_feats,
    ko_to_kegg         = FALSE,
    p_value_bar        = TRUE,
    x_lab              = "description"
  ) +
    ggprism::theme_prism() +
    ggtitle(paste(group_col)) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0)  # ← left aligned
    )

  # 8) Return objects only
  invisible(list(
    abundance          = abundance,
    metadata           = meta,
    daa_results        = daa_res,
    annotated_results  = annot,
    pval_table         = pval_table,
    top_features       = top_feats,
    errorbar_plot      = p_err,
    used_raw_p_for_plot = use_raw_p_for_plot
  ))
}


out <- run_picrust_daa_metacyc(
  file_name   = "picrust_pathways_table.tsv",
  ps_obj      = ben_phyloseq,
  group_col   = "ANY",
  recode_map  = c("0"="neg","1"="pos"),
  p_adj_cutoff = 0.15,
  p_raw_cutoff = 0.05,
  top_n        = 10
)

# View 
out$errorbar_plot        # display plot
out$used_raw_p_for_plot  # TRUE if fallback happened & if adjusted p-values are not significant

# print entire table (all rows)
pval_sorted <- out$pval_table |>
  dplyr::arrange(p_adjust, p_value)
print(pval_sorted, n = Inf)

```

